Untitled

### 2022-12-14 draft1_Rcode.Rmd
Fixed question 4 (now converges to an answer)

>>>>

```{r}
## Question 3

```{r}
## Question 4
q4.data <- read.csv("q4MixLinReg.csv",header=T)

#Part a
linRegMixEstep <- function(piCur,aCur,bCur,sCur,x,Y){
  #matrix to store E-Step values
  #row i (p(Z_{i} = 1|Y_{i}, θ^{cur}),..., p(Z_{i} = K|Y_{i}, θ^{cur})) row i has X_{i} ie X fixed
  #column j (p(Z_{i} = j|Y_{1}, θ^{cur}), ...,(Z_{i} = k|Y_{k}, θ^{cur})) column j has Z_{i}=j  ie value of Z_{i} fixed
  prob <- matrix(data=NA,nrow = length(x),ncol = length(piCur))
  for (i in 1:length(x)){
    for (j in 1:length(piCur)){
      prob[i,j] <- piCur[j]*dnorm(Y[i],mean=aCur[j]+bCur[j]*x[i],sd=sCur[j])/sum(piCur*dnorm(Y[i],mean=aCur+bCur*x[i],sd=sCur))
    }
  }
  return(prob)
}

#Part b
#i
calcNewCoefs <- function(W,x,Y){
  #Calculating a^{new) and b^{new}
  new.coef <- matrix(data=NA,ncol = 2,nrow=dim(W)[2])
  #Creating design matrix
  X <- matrix(c(rep(1,length(x)),x),ncol=2)
  for (k in 1:dim(W)[2]){
    new.coef[k,] <- solve(t(X)%*%diag(W[,k])%*%X)%*%t(X)%*%diag(W[,k])%*%Y
  }
  return(t(new.coef))
}

#ii
calcNewSd <- function(W,x,Y,a,b){
  new.sd <- NULL
  #calculating new sd
  for (k in 1:dim(W)[2]){
    new.sd[k] <- sqrt(sum(W[,k]*(Y-a[k]-b[k]*x)^2)/sum(W[,k]))
  }
  return(new.sd)
}

#iii
linRegMixMstep <- function(W,x,Y){
  #calculating new coefficents for a and b
  new.coef <- calcNewCoefs(W,x,Y)

  #calculating new sd
  new.sd <- calcNewSd(W,x,Y,new.coef[1,],new.coef[2,])

  #Calculating new pi
  new.pi <- NULL
  for (k in 1:dim(W)[2]){
    new.pi[k] <- sum(W[,k])/dim(W)[1]
  }
 return(list(new.pi,new.coef[1,],new.coef[2,],new.sd))
}

#Part c
linRegMixCalcLogLik <- function(x,Y,piCur,piNew,aCur,aNew,bCur,bNew,sCur,sNew){
  #calculating conditional probabilities for current and new values of theta
  W.cur <- linRegMixEstep(piCur,aCur,bCur,sCur,x,Y)
  W.new <- linRegMixEstep(piNew,aNew,bNew,sNew,x,Y)
  #Creating object to store values of log-likelihood for each observation i of Y 
  Q.cur <- NULL
  Q.new <- NULL
  for (i in 1:length(Y)){
    #calculating log likelihood for each value of Y
    Q.cur[i] <- sum(W.cur[i,]*log(piCur))+sum(W.cur[i,]*log(dnorm(Y[i],mean = aCur+bCur*x[i],sd=sCur)))
    Q.new[i] <- sum(W.new[i,]*log(piNew))+sum(W.new[i,]*log(dnorm(Y[i],mean = aNew+bNew*x[i],sd=sNew)))
  }
  #Summing over all observations as is in the formula
  Q.cur <- sum(Q.cur)
  Q.new <- sum(Q.new)
  return(c(Q.cur,Q.new))
}

#Part d
linRegMixEM <- function(piInit,aInit,bInit,sInit,x,Y,convergeEps){
  #Initializing the algorithm
  piCur <- piInit
  aCur <- aInit
  bCur <- bInit
  sCur <- sInit

  conv.critera <- 0
  while(conv.critera==0){
    #E-step
    W <- linRegMixEstep(piCur,aCur,bCur,sCur,x,Y)
    #Calculating new values for parameters
    theta.new <- linRegMixMstep(W,x,Y)
    #Calculating log-likelihood for current and new values of parameters
    theta.loglik <- linRegMixCalcLogLik(x,Y,piCur,theta.new[[1]],aCur,theta.new[[2]],bCur,theta.new[[3]],sCur,theta.new[[4]])
    #checking convergence criteria
    if(abs(theta.loglik[2]-theta.loglik[1])<convergeEps){
      conv.critera=1
    }
    #Updating current values of parameters
    piCur <- theta.new[[1]]
    aCur <- theta.new[[2]]
    bCur <- theta.new[[3]]
    sCur <- theta.new[[4]]
  }
  return(theta.new)
}

#Part e
mle.est <- linRegMixEM(c(0.4,0.3,0.3),c(0.1,-0.1,0.2),c(1,-1,1),c(1,0.5,1.1),q4.data$x,q4.data$Y,1e-06)
#pi MLEs
round(mle.est[[1]],3)
# (0.450, 0.255, 0.295)

#alpha MLEs
round(mle.est[[2]],3)
# (0.838,  7.458, -5.308)

#beta MLEs
round(mle.est[[3]],3)
# (0.258, -1.755,  1.223)

#sd MLEs
round(mle.est[[4]],3)
# (0.787, 0.935, 1.450)
```
